---
title: "Load and Merge Raw Crash Data"
author: "Q1 Analysis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

## Overview

This notebook loads all raw crash data files and merges them into a single comprehensive dataset for analysis.

Important: This file performs merging only. No cleaning, filtering, or feature engineering is done here. Any cleaning should be done in a separate script (e.g., `R-Q1-Ploy/01_clean_feature.R`) or another Rmd.

## Load Libraries

```{r load-libraries}
library(tidyverse)
library(janitor)
library(data.table)
library(here)
```

## 1. Load All Datasets

```{r load-data}
accident         <- fread(here("data", "accident.csv")) %>% clean_names()
accident_event   <- fread(here("data", "accident_event.csv")) %>% clean_names()
accident_location<- fread(here("data", "accident_location.csv")) %>% clean_names()
atmo             <- fread(here("data", "atmospheric_cond.csv")) %>% clean_names()
node             <- fread(here("data", "node.csv")) %>% clean_names()
person           <- fread(here("data", "person.csv")) %>% clean_names()
surface          <- fread(here("data", "road_surface_cond.csv")) %>% clean_names()
sub_dca          <- fread(here("data", "sub_dca.csv")) %>% clean_names()
vehicle          <- fread(here("data", "vehicle.csv")) %>% clean_names()
```

## 2. Check Data Dimensions

```{r check-dimensions}
data_dims <- tibble(
  Dataset = c("accident", "accident_event", "accident_location", "atmo", 
              "node", "person", "surface", "sub_dca", "vehicle"),
  Rows = c(nrow(accident), nrow(accident_event), nrow(accident_location), 
           nrow(atmo), nrow(node), nrow(person), nrow(surface), 
           nrow(sub_dca), nrow(vehicle)),
  Columns = c(ncol(accident), ncol(accident_event), ncol(accident_location), 
              ncol(atmo), ncol(node), ncol(person), ncol(surface), 
              ncol(sub_dca), ncol(vehicle))
)

print(data_dims)
```

## 3. Merge All Datasets

Keep all individual records from each table. Join tables to accident as the base table.

```{r merge-data}
merged_all <- accident %>%
    left_join(accident_event,    by = "accident_no") %>%
    left_join(accident_location,  by = "accident_no") %>%
    left_join(atmo,              by = "accident_no") %>%
    left_join(surface,           by = "accident_no") %>%
    left_join(node,              by = "accident_no") %>%
    left_join(sub_dca,           by = "accident_no") %>%
    left_join(person,            by = "accident_no") %>%
    left_join(vehicle,           by = c("accident_no", "vehicle_id"))
```

## 4. Check the Merged Dataset

```{r check-merged}
cat("Merged dataset dimensions:\n")
cat("Rows:", nrow(merged_all), "\n")
cat("Columns:", ncol(merged_all), "\n\n")

cat("First few rows:\n")
glimpse(merged_all)
```

## 5. Check for Missing Data

```{r check-missing}
missing_summary <- merged_all %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>%
  mutate(pct_missing = round(missing_count / nrow(merged_all) * 100, 2)) %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_count))

head(missing_summary, 20)
```

### Targeted missingness (vehicle-related columns only)

The overall missingness counts above include rows where vehicle columns don't apply (e.g., pedestrians or rows without a matching vehicle). The breakdown below focuses on vehicle-related rows only.

```{r check-missing-vehicle, message=FALSE}
# Helper to compute % missing that handles both character and numeric
pct_na <- function(v) {
  if (is.numeric(v)) round(mean(is.na(v)) * 100, 2) else round(mean(is.na(v) | v == "") * 100, 2)
}

# Rows that have a vehicle id (i.e., where vehicle fields logically apply)
nonveh <- merged_all %>% filter(!is.na(vehicle_id) & vehicle_id != "")

veh_missing <- tibble(
  column = c("vehicle_power", "carry_capacity", "vehicle_weight", "cubic_capacity", "vehicle_year_manuf"),
  pct_missing_vehicle_rows = c(
    pct_na(nonveh$vehicle_power),
    pct_na(nonveh$carry_capacity),
    pct_na(nonveh$vehicle_weight),
    pct_na(nonveh$cubic_capacity),
    pct_na(nonveh$vehicle_year_manuf)
  )
)

veh_missing
```

## 6. Save Merged Dataset

```{r save-data}
# Create directory if it doesn't exist
if (!dir.exists(here("data", "raw"))) {
  dir.create(here("data", "raw"), recursive = TRUE)
}

# Save merged dataset
write.csv(merged_all, here("data", "raw", "merged_all_raw.csv"), row.names = FALSE)

cat("Merged dataset saved to: data/raw/merged_all_raw.csv\n")
cat("Total rows:", nrow(merged_all), "\n")
cat("Total columns:", ncol(merged_all), "\n")
```

## Summary

- All raw data files have been loaded and merged
- The merged dataset contains all individual records (people, vehicles, events per accident)
- No filtering or summarization has been applied at this stage
- Data is saved to `data/raw/merged_all_raw.csv` for further cleaning and analysis
