---
title: "Q1 Severity Preparation"
author: "Q1 Team"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
library(tidyverse)
library(janitor)
library(data.table)
```

# Data Loading and Preparation

First, let's load our merged data:

```{r load-data}
# Load the merged dataset
merged_all <- fread("data/raw/merged_all_raw.csv")
```

# Data Cleaning for Q1 - Severity Analysis

Let's prepare our data for severity analysis:

```{r clean-severity}
# Select relevant columns for severity analysis using data.table syntax
# Note: merged_all contains raw columns; we map to expected names where needed.
q1_severity <- merged_all[, .(
    accident_no,
    severity,                 # target variable (1 Fatal, 2 Serious, 3 Minor)
    speed_zone,
    light_condition,
    weather = atmosph_cond_desc,         # from atmospheric_cond.csv
    road_surface = surface_cond_desc,    # from road_surface_cond.csv
    road_geometry_desc,
    main_vehicle_body = vehicle_body_style,
  main_vehicle_make = vehicle_make,
  age_group                            # driver/person age group
)]

# Clean speed zone data
# Treat 0, 777, 888, 999 as placeholders/unknown and set to NA
# Ensure numeric type for comparisons and summaries
q1_severity[, speed_zone := as.numeric(speed_zone)]
q1_severity[speed_zone %in% c(0, 777, 888, 999), speed_zone := NA_real_]

# Handle missing values in weather and road_surface
q1_severity[, weather := fifelse(is.na(weather), "Unknown", weather)]
q1_severity[, road_surface := fifelse(is.na(road_surface), "Unknown", road_surface)]

# Convert severity to factor with meaningful labels
q1_severity[, severity := factor(severity,
                               levels = c(1, 2, 3),
                               labels = c("Fatal", "Serious", "Minor"))]

# Handle missing speed_zone values with median imputation (based on valid speeds)
q1_severity[, speed_zone := ifelse(is.na(speed_zone),
                                  median(speed_zone, na.rm = TRUE),
                                  speed_zone)]

# Create a clean version with realistic speed limits
q1_severity_clean <- copy(q1_severity)

# Cap speeds to realistic limits (<=250 km/h)
q1_severity_clean <- q1_severity_clean[speed_zone <= 250]
```

# Data Quality Checks

Let's examine our cleaned dataset:

```{r data-checks}
# Check number of rows
cat("Number of records:", nrow(q1_severity_clean), "\n\n")

# Check speed zone distribution
cat("Speed Zone Summary:\n")
summary(q1_severity_clean$speed_zone)

# Check severity distribution
cat("\nSeverity Distribution:\n")
print(table(q1_severity_clean$severity))
```

# Summary Statistics

Let's examine our cleaned dataset:

```{r summary-stats}
# Basic summary of the cleaned data
summary(q1_severity)

# Check severity distribution
table(q1_severity$severity)
```

# Save Cleaned Dataset

```{r save-data}
# Save the cleaned dataset using data.table's fast writer
if (!dir.exists("data/processed")) dir.create("data/processed", recursive = TRUE)
fwrite(q1_severity_clean, "data/processed/q1_severity_clean.csv")
```

# Accident-level feature set (one row per accident)

We also create an accident-level dataset suitable for modeling at the crash level. This aggregates person/vehicle info per accident and avoids target leakage (we do not use per-person injury level as a predictor).

```{r accident-level}
# Helper to get first non-missing value
first_non_na <- function(x) x[which(!is.na(x) & x != "")[1]]

# Read person and vehicle raw tables to avoid duplicated counts from merged_all
dt_vehicle <- fread("data/vehicle.csv")
dt_person  <- fread("data/person.csv")

# Base accident-level features (environmental)
acc_env <- merged_all[, .(
  severity = first(severity),
  speed_zone = suppressWarnings(as.numeric(first_non_na(speed_zone))),
  light_condition = first_non_na(light_condition),
  weather = first_non_na(atmosph_cond_desc),
  road_surface = first_non_na(surface_cond_desc),
  road_geometry_desc = first_non_na(road_geometry_desc)
), by = accident_no]

# Vehicles per accident and modal vehicle attributes
veh_agg <- dt_vehicle[, .(
  vehicles_n = uniqueN(na.omit(vehicle_id)),
  main_vehicle_body = if (!all(is.na(vehicle_body_style))) names(sort(table(vehicle_body_style), decreasing = TRUE))[1] else NA_character_,
  main_vehicle_make = if (!all(is.na(vehicle_make))) names(sort(table(vehicle_make), decreasing = TRUE))[1] else NA_character_
), by = accident_no]

# Driver-only aggregates (road_user_type == 2)
drv_agg <- dt_person[road_user_type == 2, .(
  drivers_n = .N,
  driver_pct_male = mean(sex == "M", na.rm = TRUE),
  driver_age_group_mode = if (!all(is.na(age_group))) names(sort(table(age_group), decreasing = TRUE))[1] else NA_character_
), by = accident_no]

# Join all pieces
acc_df <- acc_env[veh_agg, on = "accident_no"][drv_agg, on = "accident_no"]

# Clean speed zone codes and cap
acc_df[, speed_zone := as.numeric(speed_zone)]
acc_df[speed_zone %in% c(0, 777, 888, 999), speed_zone := NA_real_]
acc_df[, speed_zone := ifelse(is.na(speed_zone), median(speed_zone, na.rm = TRUE), speed_zone)]
acc_df <- acc_df[speed_zone <= 250]

# Factor target with labels
acc_df[, severity := factor(severity, levels = c(1, 2, 3), labels = c("Fatal", "Serious", "Minor"))]

# Save accident-level dataset
fwrite(acc_df, "data/processed/q1_severity_accident.csv")
```

# Notes

- This file prepares features for Q1 severity modeling and saves `data/processed/q1_severity_clean.csv`.
- The raw merge remains in `analysis/00_load_data.Rmd`.
- Modeling is in `analysis/q1_severity_model.Rmd`.
