---
title: "Q1 Severity Modeling"
author: "Q1 Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
.libPaths("C:/Users/Pongo/Documents/R/win-library/4.5")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))

suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
})

# Use base-recommended packages for modeling to avoid heavy installs
suppressPackageStartupMessages({
  library(nnet)   # multinomial logistic regression
  library(MASS)   # ordinal logistic regression (polr)
})
```

## Overview

This notebook trains models to predict crash severity (Fatal, Serious, Minor) using the processed features from `01_q1_severity_prepare.Rmd`.

- Input: `data/processed/q1_severity_clean.csv`
- Models: Multinomial logistic (baseline), optional ordinal logistic
- Metrics: Accuracy and confusion matrix
- Output: Saved model in `models/`

## 1) Load processed data

```{r load-data}
# Prefer accident-level dataset if available; otherwise use row-level
proc_path_acc <- file.path("data", "processed", "q1_severity_accident.csv")
proc_path_row <- file.path("data", "processed", "q1_severity_clean.csv")

if (file.exists(proc_path_acc)) {
  message("Using accident-level dataset: ", proc_path_acc)
  q1 <- fread(proc_path_acc)
} else if (file.exists(proc_path_row)) {
  message("Using row-level dataset: ", proc_path_row)
  q1 <- fread(proc_path_row)
} else {
  stop("No processed dataset found. Knit analysis/00_load_data.Rmd and analysis/01_q1_severity_prepare.Rmd first.")
}

# Ensure severity is a factor in expected order
if (!is.factor(q1$severity)) {
  q1[, severity := factor(severity, levels = c("Fatal", "Serious", "Minor"))]
}

# Basic peek
print(dim(q1))
print(head(q1))
```

## 2) Minimal preprocessing for modeling

We keep the modeling set aligned with the processed features. We:
- Drop identifier columns not used for prediction (e.g., accident_no)
- Ensure character columns are factors
- Replace remaining NAs in categorical columns with "Unknown" (we already handled key fields earlier)

```{r preprocess}
# Columns available from processed step (may include):
# accident_no, severity, speed_zone, light_condition, weather, road_surface, road_geometry_desc,
# main_vehicle_body, main_vehicle_make, age_group, main_injury

# Keep a defined set of predictors if present
candidate_cols <- c(
  "accident_no", "severity", "speed_zone", "light_condition", "weather",
  "road_surface", "road_geometry_desc",
  "main_vehicle_body", "main_vehicle_make",
  "age_group",
  # accident-level features if present
  "vehicles_n", "drivers_n", "driver_pct_male", "driver_age_group_mode"
)

avail <- intersect(candidate_cols, names(q1))
q <- q1[, ..avail]

# Drop id
q[, accident_no := NULL]

# Cast character columns to factor
char_cols <- names(q)[vapply(q, is.character, logical(1))]
for (cc in char_cols) set(q, j = cc, value = factor(q[[cc]]))

# Replace NA in factor columns with "Unknown"
fac_cols <- names(q)[vapply(q, is.factor, logical(1))]
for (fc in setdiff(fac_cols, "severity")) {
  x <- q[[fc]]
  if (anyNA(x)) {
    x <- fct_explicit_na(x, na_level = "Unknown")
    set(q, j = fc, value = x)
  }
}

# Optional: ensure speed_zone is numeric
if ("speed_zone" %in% names(q)) set(q, j = "speed_zone", value = as.numeric(q$speed_zone))

# Reduce cardinality of high-cardinality categorical
if ("main_vehicle_make" %in% names(q)) {
  q[, main_vehicle_make := forcats::fct_lump_n(main_vehicle_make, n = 15)]
}

# Drop rows with missing target just in case
q <- q[!is.na(severity)]

summary(q)
```

## 3) Train/test split (stratified by severity)

```{r split}
set.seed(123)
train_idx <- q %>%
  mutate(.row = row_number()) %>%
  group_by(severity) %>%
  sample_frac(0.7) %>%
  pull(.row)

train <- q[train_idx]

test <- q[-train_idx]

cat("Train rows:", nrow(train), " Test rows:", nrow(test), "\n")
```

## 4) Baseline: majority class

```{r baseline}
maj_class <- names(sort(table(train$severity), decreasing = TRUE))[1]
preds_base <- factor(rep(maj_class, nrow(test)), levels = levels(test$severity))

acc_base <- mean(preds_base == test$severity, na.rm = TRUE)
cat("Baseline majority-class accuracy:", round(acc_base, 4), "\n")
```

## 5) Multinomial logistic regression (nnet::multinom)

```{r multinom}
# Build formula using available predictors
pred_cols <- setdiff(names(train), "severity")
form <- as.formula(paste("severity ~", paste(pred_cols, collapse = " + ")))

# To keep training time reasonable on large data, downsample training set (stratified)
max_rows <- 150000L
if (nrow(train) > max_rows) {
  set.seed(123)
  train_fit <- train %>%
    mutate(.n = row_number()) %>%
    group_by(severity) %>%
    sample_n(ceiling(max_rows * n() / nrow(train))) %>%
    ungroup() %>%
    as.data.frame()
  cat("Using stratified downsample for training:", nrow(train_fit), "rows\n")
} else {
  train_fit <- as.data.frame(train)
}

# Fit model (suppress output; raise weight cap; limit iterations)
model_multinom <- nnet::multinom(form, data = train_fit, trace = FALSE, MaxNWts = 20000, maxit = 200)

# Predict on test
preds <- predict(model_multinom, newdata = as.data.frame(test), type = "class")

# Metrics
acc <- mean(preds == test$severity, na.rm = TRUE)
cat("Multinomial accuracy:", round(acc, 4), "\n")

# Confusion matrix
print(table(Actual = test$severity, Pred = preds))
```

## 5b) Variable importance (multinomial coefficients)

```{r importance}
# Extract coefficient magnitudes across classes
coefs <- summary(model_multinom)$coefficients
if (is.list(coefs)) {
  # nnet::multinom can return list of matrices (one per non-baseline class);
  # stack them by rows so columns stay as feature names
  coef_mat <- do.call(rbind, coefs)
} else {
  coef_mat <- coefs
}

# Compute per-feature importance by summing absolute coefficients across classes
imp <- tibble(
  feature = colnames(coef_mat),
  importance = colSums(abs(coef_mat))
) %>%
  arrange(desc(importance)) %>%
  slice(1:20)

print(imp)
```

## 6) Optional: Ordinal logistic regression (MASS::polr)

```{r ordinal, eval=TRUE}
# If severity is ordinal (Minor < Serious < Fatal), we can try polr
# Note: Ensure factor order reflects increasing severity for polr
sev_ord_levels <- c("Minor", "Serious", "Fatal")
q_ord <- copy(q)
q_ord[, severity := factor(as.character(severity), levels = sev_ord_levels, ordered = TRUE)]
train_o <- q_ord[train_idx]
test_o  <- q_ord[-train_idx]

form_o <- as.formula(paste("severity ~", paste(setdiff(names(train_o), "severity"), collapse = " + ")))

polr_ok <- FALSE
model_polr <- NULL
preds_o <- NULL
acc_o <- NA_real_

try({
  model_polr <- MASS::polr(form_o, data = as.data.frame(train_o), Hess = TRUE)
  preds_o <- predict(model_polr, newdata = as.data.frame(test_o), type = "class")
  acc_o <- mean(preds_o == test_o$severity)
  polr_ok <- TRUE
}, silent = TRUE)

if (isTRUE(polr_ok)) {
  cat("Ordinal (polr) accuracy:", round(acc_o, 4), "\n")
  print(table(Actual = test_o$severity, Pred = preds_o))
} else {
  message("Skipping ordinal outputs due to fitting/prediction issue (rank deficiency or new levels).")
}
```

## 7) Save model artifacts (CSV exports)

```{r save-model}
if (!dir.exists("models")) dir.create("models", recursive = TRUE)

# --- Multinomial exports ---
# Coefficients per class
mn_coefs <- summary(model_multinom)$coefficients
if (is.list(mn_coefs)) {
  mn_coef_mat <- do.call(rbind, mn_coefs)
} else {
  mn_coef_mat <- mn_coefs
}
write.csv(mn_coef_mat, file.path("models", "q1_multinom_coefficients.csv"), row.names = TRUE)

# Confusion matrix
mn_cm <- as.matrix(table(Actual = test$severity, Pred = preds))
write.csv(mn_cm, file.path("models", "q1_multinom_confusion.csv"), row.names = TRUE)

# Metrics
mn_metrics <- data.frame(
  baseline_accuracy = if (exists("acc_base")) acc_base else NA_real_,
  multinom_accuracy = if (exists("acc")) acc else NA_real_
)
write.csv(mn_metrics, file.path("models", "q1_multinom_metrics.csv"), row.names = FALSE)

# Test-set predictions with class probabilities
mn_probs <- tryCatch(
  predict(model_multinom, newdata = as.data.frame(test), type = "probs"),
  error = function(e) NULL
)
if (!is.null(mn_probs)) {
  preds_df <- cbind(
    data.frame(Actual = test$severity, Pred = preds),
    as.data.frame(mn_probs)
  )
  write.csv(preds_df, file.path("models", "q1_multinom_test_predictions.csv"), row.names = FALSE)
}

# Variable importance (top 20)
if (exists("imp")) {
  write.csv(imp, file.path("models", "q1_multinom_importance_top20.csv"), row.names = FALSE)
}

# Training columns and target levels
write.csv(data.frame(feature = pred_cols), file.path("models", "q1_multinom_features.csv"), row.names = FALSE)
write.csv(data.frame(severity_levels = levels(q$severity)), file.path("models", "q1_severity_levels.csv"), row.names = FALSE)

# --- Ordinal (if fitted successfully) ---
if (exists("polr_ok") && isTRUE(polr_ok)) {
  plr_coefs <- coef(summary(model_polr))
  write.csv(plr_coefs, file.path("models", "q1_polr_coefficients.csv"))
  plr_cm <- as.matrix(table(Actual = test_o$severity, Pred = preds_o))
  write.csv(plr_cm, file.path("models", "q1_polr_confusion.csv"), row.names = TRUE)
  write.csv(data.frame(ordinal_accuracy = acc_o), file.path("models", "q1_polr_metrics.csv"), row.names = FALSE)
}
```

## 8) Notes

- The model trains on per-row records (merged accident-person-vehicle grain) after the cleaning step.
- For accident-level modeling, aggregate features per accident_no first in a separate script.
- Consider re-balancing (e.g., class weights or sampling) if severity classes are highly imbalanced.
